### 2017-07-13 draft/unique-linear.tex
snapshot

\begin{figure}[tp]
  \begin{align*}
    e &::= x \mid \LAM xte \mid \APP ee \mid \LET xee  \mid c \mid e+e \mid \IF eee\\
      &\mid \REC{ \overline{a=e}} \mid \GET ea \mid \UPD eae \\
    O,P,Q &::= \NOTHING \mid \READ \mid \WRITE & \NOTHING & \sqsubseteq\WRITE\sqsubset \READ \\
    s,t &::= \NUM \mid s \to t \mid \REC{ \overline{a :^P t}} \\
  \end{align*}
  \caption{Syntax}
  \label{fig:syntax-3}
\end{figure}
In the syntax in Figure~\ref{fig:syntax-3}, effects are attached to single record labels, intuitively $P, Q \subseteq
\{ \READ, \WRITE \}$ but it does not make sense to distinguish
$\{\WRITE, \READ\}$ from $\{\WRITE\}$, so we take the three element lattice as indicated.

@@

@@

  \inferrule
  { A_1 \vdash e_1 : t_1 \\
    A_2, x : t_1 \vdash e_2 : t_2 \\
    A = A_1 \OVERRIDE A_2
  }
  { A \vdash \LET x{e_1} e_2 : t}
  \end{mathpar}
  \caption{Typing rules}
  \label{fig:field-effects}
\end{figure}
The novel ingredient in the typing rules are the propagation and splitting rules for the effect annotations on
record labels.
Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
\begin{itemize}
\item $P = \NOTHING$ propagate $Q=\NOTHING$.
\item $P = \READ$ propagate $Q = \WRITE$, no action required inside closure.
\item $P = \WRITE$ propagate $Q = \WRITE$, perform copy inside of closure.
\end{itemize}
 Propagating $\READ$ is not safe in both cases because that enables a write operation on the same
 record after the closure has been constructed. Alternatively, we could have another READONLY mode,
 which prevents subsequent write operations.

@@

  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a\sqcup\READ &\WRITE& \NOTHING & \WRITE & P_a & \WRITE
                           & \text{copy, propagate left to write} \\
    \READ & \WRITE& O_a & \WRITE & \NOTHING & \WRITE
                           & \text{copy, propagate left to write} \\
    P_a\sqcup\READ   &\WRITE& \READ & \WRITE & P_a & \WRITE
                           &  \text{copy, propagate left to write}\\
    \READ&\WRITE& \WRITE & \WRITE & \READ & \WRITE
                       &  \text{copy, propagate left to write}\\
    \WRITE &\WRITE& \WRITE & \WRITE & \WRITE & \WRITE
                       &\text{copy, propagate left to write} \\
  \end{array}
\end{mathpar}
A copy operation reads \textbf{every} field. Hence, a copy operation
caused by one field implies a read effect on all fields.

\end{document}

>>>>

\begin{figure}[tp]
  \begin{align*}
    e,f,g,r
    &::= x \mid \LAM xte \mid \APP fe \mid \FIX xtf \\
    & \mid \LET xef  \mid c \mid f+g \mid \IF efg\\
    &\mid \REC{ \overline{a=e}} \mid \GET ra \mid \UPD rae \\
    O,P,Q &::= \PVAR \mid \NOTHING \mid \READ \mid \WRITE & \NOTHING & \sqsubseteq\WRITE\sqsubset \READ \\
    s,t &::= \TVAR \mid \NUM \mid s \to t \mid \REC{ \overline{a :^P t}} \\
    S, T& ::= \forall \TVAR. T \mid \forall \PVAR. T \mid t
  \end{align*}
  \caption{Syntax}
  \label{fig:syntax-3}
\end{figure}
In the syntax in Figure~\ref{fig:syntax-3}, effects are attached to single record labels, intuitively $P, Q \subseteq
\{ \READ, \WRITE \}$ but it does not make sense to distinguish
$\{\WRITE, \READ\}$ from $\{\WRITE\}$, so we take the three element lattice as indicated.

@@

  \inferrule[PT: Subtyping needed?]
  { A, x:s \vdash f :t  \\ s \SUB t}
  { A \vdash \FIX xtf : t}

@@

  \inferrule
  { A_1 \vdash e_1 : t_1 \\
    A_2, x : t_1 \vdash e_2 : t_2 \\
    A = A_1 \OVERRIDE A_2
  }
  { A \vdash \LET x{e_1} e_2 : t}

  \inferrule
  {
    A = A_1 \OVERRIDE (A_2 \sqcup A_3) \\
    A_1 \vdash e : \NUM \\
    A_2 \vdash f : t_2 \\
    A_3 \vdash g : t_3 \\
    \SPLIT{t_1}{  t_2}{ t_3}
  }
  {
    A \vdash \IF e f g : t_1
  }
  % PT: is subtyping really needed? We are propagating demand information *backwards*, so (e.g.) the demand
  % on the branches of a conditional is the same as the demand on the result of the conditional.
  % \inferrule[Sub]
  % {A \vdash e : s \\ s \SUB t}
  % {A \vdash e : t}
  \end{mathpar}
  \caption{Typing rules.
  It seems like the subtyping in the FIX rule can alway be eliminated by unfolding the fixed point
  once.
  If the demand on $t$ is stronger or weaker than $s$, then unfolding will mediate the demand.
  NB the directionality of subtyping is wrong if interpolated from the lattice ordering (as it is presently).
}
  \label{fig:field-effects}
\end{figure}
% \begin{figure}[tp]
%   \begin{mathpar}
%     \inferrule{}{\NUM \SUB \NUM}
%
%     \inferrule{
%       s' \SUB s \\ t \SUB t'
%     }{
%       s \to t \SUB s' \to t' 
%     }
%
%     \inferrule{
%       \overline{s_i \SUB t_i} \\
%       \overline{P_i \sqsubseteq Q_i}
%     }{
%       \REC{ \overline{a_i :^{P_i} s_i}}
%       \SUB
%       \REC{ \overline{a_i :^{Q_i} t_i}}
%     }
%   \end{mathpar}
%   \caption{Subtyping}
%   \label{fig:subtyping-3}
% \end{figure}
\begin{figure}[tp]
  \begin{align*}
    \NUM \sqcup \NUM
    & = \NUM
    \\
    (s \to t) \sqcup (s' \to t')
    & = (s \sqcap s') \to (t \sqcup t') \qquad ???
    \\
    \REC{ \overline{a_i :^{P_i} s_i}} \sqcup \REC{ \overline{a_i :^{Q_i} t_i}}
    &= \REC{ \overline{a_i :^{P_i \sqcup Q_i} (s_i \sqcup t_i)}}
  \end{align*}
  \begin{mathpar}
    \inferrule{}{
      \SPLIT \NUM \NUM \NUM
    }

    \inferrule{}{
      \JOIN \NUM \NUM \NUM
    }
    \\
    \inferrule{
      \JOIN{s}{ s_1}{ s_2} \\
      \SPLIT{t}{t_1}{t_2}
    }{
      \SPLIT{s \to t}{s_1 \to t_1}{s_2 \to t_2}
    }

    \inferrule{
      ??? \\
      \JOIN{s}{ s_1}{ s_2} \\
      \SPLIT{t}{t_1}{t_2}
    }{
      \JOIN{s \to t}{s_1 \to t_1}{s_2 \to t_2}
    }
    \\
    \inferrule
    {\overline{\SPLIT{s_i}{s_i'}{s_i''}}}
    {\SPLIT{\REC{ \overline{a_i :^{P_i} s_i}}}{\REC{ \overline{a_i :^{P_i} s_i'}}}{\REC{ \overline{a_i :^{P_i} s_i''}}}}
  \end{mathpar}
  \caption{Bounding and splitting operations on types}
  \label{fig:bounding-3}
\end{figure}
The novel ingredient in the typing rules are the propagation and splitting rules for the effect annotations on
record labels indicated by the ternary override relation $\OVERRIDE$.
Function closures have ``infinite'' splitting $(a :^P t)^*$ into $(a :^Q t)$.
\begin{itemize}
\item $P = \NOTHING$ propagate $Q=\NOTHING$.
\item $P = \READ$ propagate $Q = \WRITE$, no action required inside closure.
\item $P = \WRITE$ propagate $Q = \WRITE$, perform copy inside of closure.
\end{itemize}
 Propagating $\READ$ is not safe in both cases because that enables a write operation on the same
 record after the closure has been constructed. Alternatively, we could have another READONLY mode,
 which prevents subsequent write operations.

@@

  \begin{array}{|c|c||c|c||c|c|l}
    Q_a&Q_b&O_a&O_b&P_a&P_b& \\
    \hline
    P_a\sqcup\READ &\WRITE& \NOTHING & \WRITE & P_a & \WRITE
                           & \text{copy, propagate left to write} \\
    \READ & \WRITE& O_a & \WRITE & \NOTHING & \WRITE
                           & \text{copy, propagate left to write} \\
    P_a\sqcup\READ   &\WRITE& \READ & \WRITE & P_a & \WRITE
                           &  \text{copy, propagate left to write}\\
    \READ&\WRITE& \WRITE & \WRITE & \READ & \WRITE
                       &  \text{copy, propagate left to write}\\
    \WRITE &\WRITE& \WRITE & \WRITE & \WRITE & \WRITE
                       &\text{copy, propagate left to write} \\
  \end{array}
\end{mathpar}
A copy operation reads \textbf{every} field. Hence, a copy operation
caused by one field implies a read effect on all fields.
\clearpage{}
Consider the typing for 
\begin{align*}
  f
  &=(\LAM r{\REC{ a:^\WRITE \NUM}} \UPD r a{42})
  &&: \REC{ a:^\WRITE \NUM} \to \REC{ a:^\PVAR \NUM}
  \\
  g
  &=(\LAM r{\REC{a :^\NOTHING \NUM}} \REC{a = 0})
  &&: \REC{a :^\NOTHING \NUM} \to \REC{ a:^\PVAR \NUM}
  \\
  &\IF e{f}{g}
  &&: \REC{ a:^\WRITE \NUM} \to \REC{ a:^\PVAR \NUM}
\end{align*}
This propagation is correct,
\textbf{but} if the write is propagated, this may cause a useless copy if $e$ is false.
\end{document}
